---
title: "STAT5003 Assignment -w7"
author: "Jeannie Chang, Zihan Li"
output:
  html_document:
    code_folding: hide 
---

## 0. Introduction

Australia’s population is constantly shifting. Some regions are ageing as younger residents move away, while others are growing younger due to new housing developments and family migration. For infrastructure planners, predicting these shifts is essential to ensure services like schools, hospitals, and aged care are built where they will be needed most.

This project uses Australian Bureau of Statistics (ABS) data to classify SA2 regions by their "demographic trajectory". We aim to identify which socio-economic features best predict whether a community’s profile is changing, providing a tool for proactive regional planning.

## 1. Framing & Dataset Selection

### 1-1 Research Question

How accurately can we predict if an Australian region (SA2) is ageing, stable, or growing younger based on its current socio-economic features, and what does this tell planners about future service needs?

### 1-2 Project Methodology

We use a two-step approach to solve the classification task:
- *Label Creation (Calculated)*: We calculate the change in median age from 2011 to 2025 using ABS longitudinal data. We then group regions into three categories: Ageing, Stable, or Growing Younger.
- *Classification (Predictive)*: We build five supervised models (e.g., k-NN, Random Forest) to predict these labels using current income and population density data.

## 2. Consider & Gather Data

### 2-1 Data Sources

We use two spreadsheets from the ABS *Data by Region, 2011–2025* release (https://www.abs.gov.au/methodologies/data-region-methodology/2011-24#data-downloads):

- **`14100DO0001_2011-25(1).xlsx`** — regional population and demographic variables (age structure, sex composition, population density, working-age population, etc.)
- **`14100DO0004_2011-25(1).xlsx`** — regional income variables (median income, income bracket distributions, etc.)

The longitudinal structure is essential: we derive trajectory labels from year-over-year change in demographic indicators, then use income and population features to predict those labels.

The two datasets pass the CRAAP quality test:
- **Currency & Relevance**: Latest ABS Data by Region 2011–2025 release; ideal for measuring demographic change.
- **Authority & Accuracy**: Sourced from the Australian Bureau of Statistics (ABS), the national standard for Australian demographic data.
- **Purpose**: Objective, public-research data designed for regional analysis and policy planning.

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(scales)
library(knitr)

# ── File paths ──────────────────────────────────────────────────────────────
population_file <- "14100DO0001_2011-25.xlsx"
income_file     <- "14100DO0004_2011-25.xlsx"

# ── 1. Sheet names ──────────────────────────────────────────────────────────
pop_sheets    <- excel_sheets(population_file)
income_sheets <- excel_sheets(income_file)

cat("Population file ->", length(pop_sheets), "sheets :", paste(pop_sheets, collapse = ", "), "\n")
cat("Income file     ->", length(income_sheets), "sheets :", paste(income_sheets, collapse = ", "), "\n")

# ── 2. Shape of every Table sheet ──────────────────────────────────────────
pop_table_sheets    <- pop_sheets[str_detect(pop_sheets, "^Table")]
income_table_sheets <- income_sheets[str_detect(income_sheets, "^Table")]

cat("\n── Population file ──────────────────────────────────────\n")
for (sheet in pop_table_sheets) {
  tmp <- read_excel(population_file, sheet = sheet, skip = 6)
  cat(sprintf("  %-12s: %6s rows  x  %4s columns\n", sheet, format(nrow(tmp), big.mark = ","), ncol(tmp)))
}

cat("\n── Income file ──────────────────────────────────────────\n")
for (sheet in income_table_sheets) {
  tmp <- read_excel(income_file, sheet = sheet, skip = 6)
  cat(sprintf("  %-12s: %6s rows  x  %4s columns\n", sheet, format(nrow(tmp), big.mark = ","), ncol(tmp)))
}

# ── 3. Load and combine all Table sheets ───────────────────────────────────
# not clean here because we need to calculate the dirty rate
load_all_tables <- function(filepath, table_sheets) {
  map_dfr(table_sheets, function(sheet) {
    read_excel(
      filepath,
      sheet = sheet,
      skip = 6,
      na = c("-", "..", "", " ", "np", "n.p.", "–", "N/A"),
      .name_repair = "unique"
    ) %>%
      mutate(
        source = sheet,
        t2 = sheet == "Table 2",
        t3 = sheet == "Table 3"
      )
  })
}

population_raw <- load_all_tables(population_file, pop_table_sheets)
income_raw     <- load_all_tables(income_file, income_table_sheets)
```

### 2-2 Dataset Justification

The dataset meets the project requirements as follows:

```{r}
# ── Create popul_df and income_df ──────────────────────────────────────────
popul_df  <- population_raw
income_df <- income_raw

# Identify non-identifier columns
id_cols       <- c("Code", "Label", "Year", "source", "t2", "t3")
data_cols_pop <- setdiff(names(popul_df), id_cols)
data_cols_inc <- setdiff(names(income_df), id_cols)

# Convert data columns to numeric
popul_df  <- popul_df %>% mutate(across(all_of(data_cols_pop), ~ suppressWarnings(as.numeric(.x))))
income_df <- income_df %>% mutate(across(all_of(data_cols_inc), ~ suppressWarnings(as.numeric(.x))))

# Outlier rate: Tukey IQR fence
iqr_outlier_rate <- function(df, cols) {
  vals <- as.matrix(df[, cols, drop = FALSE])
  q1 <- apply(vals, 2, quantile, probs = 0.25, na.rm = TRUE)
  q3 <- apply(vals, 2, quantile, probs = 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower <- q1 - 1.5 * iqr
  upper <- q3 + 1.5 * iqr

  valid <- !is.na(vals)
  is_out <- sweep(vals, 2, lower, `<`) | sweep(vals, 2, upper, `>`)
  is_out[!valid] <- FALSE

  if (sum(valid) == 0) return(NA_real_)
  sum(is_out) / sum(valid)
}

pop_outlier <- iqr_outlier_rate(popul_df, data_cols_pop)
inc_outlier <- iqr_outlier_rate(income_df, data_cols_inc)

# ── Summary table ───────────────────────────────────────────────────────────
pop_miss <- mean(is.na(as.matrix(popul_df[, data_cols_pop, drop = FALSE])))
inc_miss <- mean(is.na(as.matrix(income_df[, data_cols_inc, drop = FALSE])))

summary_tbl <- tibble(
  DataFrame = c("popul_df", "income_df"),
  Rows = c(format(nrow(popul_df), big.mark = ","), format(nrow(income_df), big.mark = ",")),
  `Total columns` = c(ncol(popul_df), ncol(income_df)),
  `Data columns` = c(length(data_cols_pop), length(data_cols_inc)),
  `Overall missing %` = c(percent(pop_miss, accuracy = 0.1), percent(inc_miss, accuracy = 0.1)),
  `Potential outliers %` = c(percent(pop_outlier, accuracy = 0.1), percent(inc_outlier, accuracy = 0.1))
)

knitr::kable(summary_tbl)
```

4 requirements met:
- **Large**: 26k+ population rows and 32k+ income rows across all geographic levels; SA2-level alone exceeds 10,000 observations.
- **Messy**: >40% overall missing value rate in both files (ABS suppresses small counts), plus a meaningful share of potential outliers in non-missing values.
- **Complex**: 163 population columns and 56 income columns combined (>100 threshold).
- **Integration**: data is from two source.

We also note that rows span multiple geographic levels (National, State, SA4, SA3, LGA, SA2), which overlap hierarchically. For our trajectory analysis we retain only SA2 rows, as this is the finest level with consistent longitudinal coverage.

```{r}
# ── Geographic level breakdown: row count + missing rate ───────────────────
classify_geo <- function(code, t2, t3) {
  s <- str_trim(as.character(code))

  case_when(
    s == "AUS" ~ "National",
    str_detect(s, "^[0-9]$") ~ "State/Territory",
    str_detect(s, "^[0-9]{3}$") & t3 ~ "Remote",
    str_detect(s, "^[0-9]{3}$") ~ "SA4",
    nchar(s) == 5 & !str_detect(s, "^[0-9]+$") ~ "GCCSA",
    str_detect(s, "^[0-9]{5}$") & t2 ~ "LGA",
    str_detect(s, "^[0-9]{5}$") & !t2 ~ "SA3",
    str_detect(s, "^[0-9]{9}$") ~ "SA2",
    TRUE ~ "Other"
  )
}

popul_df <- popul_df %>% mutate(geo_level = classify_geo(Code, t2, t3))
income_df <- income_df %>% mutate(geo_level = classify_geo(Code, t2, t3))

pop_geo_counts <- popul_df %>% count(geo_level, name = "rows")
income_geo_counts <- income_df %>% count(geo_level, name = "rows")

pop_missing <- popul_df %>%
  group_by(geo_level) %>%
  summarise(missing_rate = mean(is.na(as.matrix(across(all_of(data_cols_pop))))), .groups = "drop")

inc_missing <- income_df %>%
  group_by(geo_level) %>%
  summarise(missing_rate = mean(is.na(as.matrix(across(all_of(data_cols_inc))))), .groups = "drop")

geo_order <- c("National", "State/Territory", "GCCSA", "SA4", "SA3", "LGA", "SA2")

row_df <- bind_rows(
  pop_geo_counts %>% mutate(file = "Population"),
  income_geo_counts %>% mutate(file = "Income")
) %>%
  mutate(geo_level = factor(geo_level, levels = geo_order))

miss_df <- bind_rows(
  pop_missing %>% mutate(file = "Population"),
  inc_missing %>% mutate(file = "Income")
) %>%
  mutate(geo_level = factor(geo_level, levels = geo_order))

ggplot(row_df, aes(x = geo_level, y = rows, fill = file)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Population" = "steelblue", "Income" = "coral")) +
  labs(
    title = "Row count by geographic level (evidence: Large criterion)",
    x = "Geographic level",
    y = "Number of rows"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

ggplot(miss_df, aes(x = geo_level, y = missing_rate, fill = file)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Population" = "steelblue", "Income" = "coral")) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Missing value rate by geographic level (evidence: Messy criterion)",
    x = "Geographic level",
    y = "Missing rate"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

SA2 is the preferred geographic level for this project: it provides sufficient granularity to capture local demographic change while having enough observations (>10,000 labelled rows after year-on-year differencing) for reliable classification. We filter both cleaned dataframes to SA2 and carry forward only these rows for all subsequent steps.

```{r}
# Filter to SA2 level using geo_level column assigned above
population_sa2 <- popul_df %>% filter(geo_level == "SA2")
income_sa2     <- income_df %>% filter(geo_level == "SA2")

cat("population_sa2 :", format(nrow(population_sa2), big.mark = ","), "rows\n")
cat("income_sa2     :", format(nrow(income_sa2), big.mark = ","), "rows\n")
```



### 2-3 Cleaning Strategy and Merge Preparation

After limiting the two files to the SA2 level, the next step is to prepare them for integration, rather than repeating the previous descriptive summary. At this stage, we mainly focus on three issues: standardizing bonding, applying transparent missing value strategy, and checking whether the generated SA2 year structure is suitable for future tag construction.

The ABS file contains non-standard missing value tags, such as' - `, `. `, `np` and blank cells. These have been converted to "Na" during the import process. However, not all deficiencies should be treated in the same way. Consistent with the project plan, we use **key variables** -- in this case, the column whose name contains "total" - to evaluate whether a row is reliable enough. Delete too many missing rows in the key variable while retaining the remaining missing values until merged, and then deal with them in a more conservative manner.


```{r}
# ===== Prepare SA2 data for integration =====
pop_merge_prep <- population_sa2 %>%
  mutate(
    code = str_trim(as.character(Code)),
    year = as.integer(Year)
  )

inc_merge_prep <- income_sa2 %>%
  mutate(
    code = str_trim(as.character(Code)),
    year = as.integer(Year)
  )
```

Since the final analysis unit in this project is **SA2 year**, the merge key must be standardized before adding. Therefore, we create clean "code" and "year" fields in both datasets, and then check whether each dataset contains a unique "code year" combination. This is important because duplicate keyframes create many to many connections and distort downstream results.

```{r}
# ===== Check whether code-year is unique in each dataset =====
pop_key_check <- pop_merge_prep %>%
  count(code, year, name = "n") %>%
  filter(n > 1)

inc_key_check <- inc_merge_prep %>%
  count(code, year, name = "n") %>%
  filter(n > 1)

cat("Population duplicate code-year rows:", nrow(pop_key_check), "
")
cat("Income duplicate code-year rows:", nrow(inc_key_check), "
")
```

According to the project annotation, then apply the missing value rule based on the key variable. We identify the variable containing "total" as a summary indicator of whether the row contains available information. Rows with more than 50% missing values in these key columns are considered unreliable and deleted. For the remaining rows, values are not deleted simply because the two datasets are not aligned exactly. Instead, they are saved after integration so that potentially useful SA2 year observations are not lost prematurely.

```{r}
# ===== Key-based missing value treatment before integration =====
key_cols_pop <- names(pop_merge_prep)[
  str_detect(names(pop_merge_prep), regex("total", ignore_case = TRUE)) &
    !(names(pop_merge_prep) %in% c("Code", "Label", "Year", "source", "t2", "t3", "geo_level", "code", "year"))
]

key_cols_inc <- names(inc_merge_prep)[
  str_detect(names(inc_merge_prep), regex("total", ignore_case = TRUE)) &
    !(names(inc_merge_prep) %in% c("Code", "Label", "Year", "source", "t2", "t3", "geo_level", "code", "year"))
]

pop_merge_ready <- pop_merge_prep %>%
  mutate(missing_ratio_key = if (length(key_cols_pop) > 0) rowMeans(is.na(across(all_of(key_cols_pop)))) else 0) %>%
  filter(missing_ratio_key <= 0.5)

inc_merge_ready <- inc_merge_prep %>%
  mutate(missing_ratio_key = if (length(key_cols_inc) > 0) rowMeans(is.na(across(all_of(key_cols_inc)))) else 0) %>%
  filter(missing_ratio_key <= 0.5)

cat("Population SA2 after key-column filtering:", format(nrow(pop_merge_ready), big.mark = ","), "rows
")
cat("Income SA2 after key-column filtering:", format(nrow(inc_merge_ready), big.mark = ","), "rows
")
```

This strategy strictly follows the project logic: we do **not** delete observations simply because they cannot match immediately between files. Instead, we delete only the substantively unreliable rows based on the missing in the key variable. This preserves the overall vertical structure, which will be used later to build the result tag.

### 2-4 Evidence for Integration Feasibility

The last task in this section is to demonstrate that two cleaned SA2 datasets can be effectively integrated into a single analysis file. This is the strongest evidence of the **integration** standard in the guidelines: combine multiple tables in each source file, limit the two sources to a common geographical unit, and then link the two cleaned SA2 panels to the shared key.

Although it was initially thought that the complete outer join can retain all possible records in the two datasets, the **left join** is more suitable for the current project goal. This is because the classification label will be derived from the population change in the population file later. Therefore, the population SA2 year team has determined the core analysis structure, and the income variable is attached as the prediction feature where the matching record is available.

```{r}
# ===== Final merge at SA2 year level =====
sa2_merged <- left_join(
  pop_merge_ready,
  inc_merge_ready,
  by = c("code", "year"),
  suffix = c("_pop", "_inc")
)

cat(
  "Merged SA2 year dataset:",
  format(nrow(sa2_merged), big.mark = ","), "rows x", ncol(sa2_merged), "columns
"
)
```

Using the left join means that mismatched revenue records will not result in discarding valid overall observations. On the contrary, mismatched feature fields are retained as missing values and can be transparently processed in subsequent preprocessing. This is preferable to the internal connection, because the internal connection will delete the effective overall observation value of SA2 year and may lead to deviation in the later label construction.

To assess the integration quality, we check how many lines in the consolidated SA2 year file successfully received at least one revenue side variable.

```{r}
# ===== Merge quality assessment =====
income_only_cols <- setdiff(names(inc_merge_ready), names(pop_merge_ready))
income_only_cols <- setdiff(income_only_cols, c("code", "year", "missing_ratio_key"))

if (length(income_only_cols) > 0) {
  merge_quality <- sa2_merged %>%
    summarise(
      total_rows = n(),
      matched_income_rows = sum(rowSums(!is.na(across(all_of(income_only_cols)))) > 0),
      unmatched_income_rows = sum(rowSums(!is.na(across(all_of(income_only_cols)))) == 0)
    )
  knitr::kable(merge_quality)
}
```

In general, this section shows that the selected ABS data is not only huge, messy and complex, but also really suitable for integration. Merge multiple tables in each file, use the defensible key variable rules to deal with the missing, and merge the two cleaned up SA2 data sets into a unified SA2 year structure to prepare for the subsequent exploratory analysis and label creation.

## 3. Exploratory Data Analysis (EDA)

### 3-1 Univariate Analysis

The high overall missing rate (~40–65%) observed in Section 2 partially reflects ABS suppression of small counts: a value is withheld when it is too small to publish, not because it is truly absent. For percentage breakdown columns (age group shares, income bracket shares), a suppressed cell effectively means the proportion rounds to zero.

We therefore distinguish two types of column:
- **Key columns** — those whose name contains *"total"* — which serve as summary measures and should be populated for all valid SA2 observations.
- **Breakdown columns** — all others — where NaN can legitimately represent zero.

The approach: identify key columns, compute their missing rate, and drop rows where all key columns are NaN (malformed rows). Then impute 0 into remaining NaN cells across breakdown columns.

Before detailed EDA, the long ABS column names are shortened into readable snake-case names. This does not change the data itself; it creates a dictionary that makes later plots and modelling code easier to read. The original names are still retained in `name_dictionary` for traceability.

```{r}
# ---- Short readable variable-name dictionary ----
shorten_abs_name <- function(x) {
  x %>%
    str_to_lower() %>%
    str_replace_all("estimated resident population", "erp") %>%
    str_replace_all("population density", "pop_density") %>%
    str_replace_all("working age population", "working_age_pop") %>%
    str_replace_all("median age", "median_age") %>%
    str_replace_all("median equivalised total household income", "median_eq_hh_income") %>%
    str_replace_all("currently serving in the australian defence force", "serving_defence_force") %>%
    str_replace_all("males", "male") %>%
    str_replace_all("females", "female") %>%
    str_replace_all("persons", "person") %>%
    str_replace_all("\\(no\\.\\)", "n") %>%
    str_replace_all("\\(\\%\\)", "pct") %>%
    str_replace_all("\\(years\\)", "yrs") %>%
    str_replace_all("\\(weekly\\)", "weekly") %>%
    str_replace_all("\\$", "") %>%
    str_replace_all("[^a-z0-9]+", "_") %>%
    str_replace_all("^_|_$", "") %>%
    str_replace_all("_+", "_") %>%
    str_replace("^([0-9])", "v_\\1")
}

data_cols_for_dictionary <- unique(c(
  setdiff(names(population_sa2), c("Code", "Label", "Year", "source", "t2", "t3", "geo_level")),
  setdiff(names(income_sa2), c("Code", "Label", "Year", "source", "t2", "t3", "geo_level"))
))

name_dictionary <- tibble(
  original_name = data_cols_for_dictionary,
  short_name = make.unique(shorten_abs_name(data_cols_for_dictionary), sep = "_dup")
)

knitr::kable(head(name_dictionary, 15))
```

```{r, fig.width=12, fig.height=8}
# ── Non-missing rate per column ─────────────────────────────────────────────
data_cols_sa2_pop_all <- setdiff(names(population_sa2), c("Code", "Label", "Year", "source", "t2", "t3", "geo_level"))
data_cols_sa2_inc_all <- setdiff(names(income_sa2), c("Code", "Label", "Year", "source", "t2", "t3", "geo_level"))

pop_nonmiss <- sort(colMeans(!is.na(as.matrix(population_sa2[, data_cols_sa2_pop_all, drop = FALSE]))), decreasing = TRUE)
inc_nonmiss <- sort(colMeans(!is.na(as.matrix(income_sa2[, data_cols_sa2_inc_all, drop = FALSE]))), decreasing = TRUE)

make_panel_df <- function(x, dataset_name) {
  top10 <- tibble(
    column = names(head(x, 10)),
    non_missing_rate = as.numeric(head(x, 10)),
    panel = paste(dataset_name, "— Top 10 most complete columns")
  )

  bottom10 <- tibble(
    column = names(rev(tail(x, 10))),
    non_missing_rate = as.numeric(rev(tail(x, 10))),
    panel = paste(dataset_name, "— Bottom 10 least complete columns")
  )

  bind_rows(top10, bottom10)
}

nonmiss_plot_df <- bind_rows(
  make_panel_df(pop_nonmiss, "population_sa2"),
  make_panel_df(inc_nonmiss, "income_sa2")
) %>%
  mutate(column = str_trunc(column, 55))

ggplot(nonmiss_plot_df, aes(x = reorder(column, non_missing_rate), y = non_missing_rate, fill = panel)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ panel, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Non-Missing Value Rate by Column (SA2 level)",
    x = NULL,
    y = "Non-missing rate"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    strip.text = element_text(size = 10, face = "bold"),
    panel.spacing = unit(1.1, "lines")
  )
```

We select only column names containing *"total"* as key columns and compute their missing rate. Rows where all key columns are NaN are considered malformed and dropped. This ensures downstream analysis is not distorted by rows that contain no usable summary data.

```{r}
# ── Key columns: names containing 'total' ──────────────────────────────────
id_cols_full <- c("Code", "Label", "Year", "source", "t2", "t3", "geo_level")

key_cols_pop <- names(population_sa2)[str_detect(str_to_lower(names(population_sa2)), "total") & !(names(population_sa2) %in% id_cols_full)]
key_cols_inc <- names(income_sa2)[str_detect(str_to_lower(names(income_sa2)), "total") & !(names(income_sa2) %in% id_cols_full)]

cat("Key population columns (", length(key_cols_pop), "):\n", sep = "")
cat(paste0("  ", key_cols_pop, collapse = "\n"), "\n")
cat("\nKey income columns (", length(key_cols_inc), "):\n", sep = "")
cat(paste0("  ", key_cols_inc, collapse = "\n"), "\n")

# Missing rate on key columns only
cat("\n── Missing rate on key columns ─────────────────────────\n")
cat("\nPopulation:\n")
print(percent(colMeans(is.na(as.matrix(population_sa2[, key_cols_pop, drop = FALSE]))), accuracy = 0.1))
cat("\nIncome:\n")
print(percent(colMeans(is.na(as.matrix(income_sa2[, key_cols_inc, drop = FALSE]))), accuracy = 0.1))

# Drop rows where ALL key columns are NA
population_sa2_clean <- population_sa2 %>%
  filter(rowSums(!is.na(across(all_of(key_cols_pop)))) > 0)

income_sa2_clean <- income_sa2 %>%
  filter(rowSums(!is.na(across(all_of(key_cols_inc)))) > 0)

cat(
  "\nPopulation SA2 :", format(nrow(population_sa2), big.mark = ","), "→",
  format(nrow(population_sa2_clean), big.mark = ","), "rows  (dropped",
  format(nrow(population_sa2) - nrow(population_sa2_clean), big.mark = ","), ")\n"
)
cat(
  "Income SA2     :", format(nrow(income_sa2), big.mark = ","), "→",
  format(nrow(income_sa2_clean), big.mark = ","), "rows  (dropped",
  format(nrow(income_sa2) - nrow(income_sa2_clean), big.mark = ","), ")\n"
)
```

After deleting the lines with abnormal format or missing core information, the remaining Nan value in the sub category variable will be filled as 0. It is reasonable to do so, because ABS will hide data whose value is too small for public release; For our percentage variable, this kind of hidden small value can basically be regarded as a proportion close to 0.

```{r}
# ── Impute NA → 0 in breakdown data columns ────────────────────────────────
id_cols_full <- c("Code", "Label", "Year", "source", "t2", "t3", "geo_level")
data_cols_sa2_pop <- setdiff(names(population_sa2_clean), id_cols_full)
data_cols_sa2_inc <- setdiff(names(income_sa2_clean), id_cols_full)

population_sa2_clean <- population_sa2_clean %>%
  mutate(across(all_of(data_cols_sa2_pop), ~ replace_na(.x, 0)))

income_sa2_clean <- income_sa2_clean %>%
  mutate(across(all_of(data_cols_sa2_inc), ~ replace_na(.x, 0)))

cat("Remaining NA after imputation:\n")
cat("  population_sa2_clean :", sum(is.na(as.matrix(population_sa2_clean[, data_cols_sa2_pop, drop = FALSE]))), "\n")
cat("  income_sa2_clean     :", sum(is.na(as.matrix(income_sa2_clean[, data_cols_sa2_inc, drop = FALSE]))), "\n")
```

For outlier detection, we focus on the key column, rather than the subdivision of each small percentage. Extreme SA2 values may be real, such as dense downtown areas or very small remote areas. Therefore, the Tukey IQR rule is used to mark and limit outliers to model ready replicas rather than automatically deleting entire rows.

```{r}
# ---- Outlier detection and conservative treatment on key columns ----
iqr_bounds <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  c(lower = q1 - 1.5 * iqr, upper = q3 + 1.5 * iqr)
}

winsorise_iqr <- function(x) {
  if (all(is.na(x))) return(x)
  b <- iqr_bounds(x)
  pmin(pmax(x, b["lower"]), b["upper"])
}

flag_outliers <- function(df, cols) {
  map_dfr(cols, function(col) {
    b <- iqr_bounds(df[[col]])
    tibble(
      variable = col,
      lower_bound = b["lower"],
      upper_bound = b["upper"],
      outlier_rows = sum(df[[col]] < b["lower"] | df[[col]] > b["upper"], na.rm = TRUE),
      outlier_rate = mean(df[[col]] < b["lower"] | df[[col]] > b["upper"], na.rm = TRUE)
    )
  })
}

pop_outlier_tbl <- flag_outliers(population_sa2_clean, key_cols_pop)
inc_outlier_tbl <- flag_outliers(income_sa2_clean, key_cols_inc)

outlier_display_tbl <- bind_rows(
  pop_outlier_tbl %>% mutate(file = "Population"),
  inc_outlier_tbl %>% mutate(file = "Income")
) %>%
  arrange(desc(outlier_rate)) %>%
  mutate(
    variable = str_trunc(variable, 65),
    outlier_rate = percent(outlier_rate, accuracy = 0.1),
    lower_bound = round(lower_bound, 2),
    upper_bound = round(upper_bound, 2)
  ) %>%
  select(file, variable, outlier_rows, outlier_rate, lower_bound, upper_bound)

knitr::kable(
  head(outlier_display_tbl, 15),
  caption = "Top 15 key variables with the highest potential outlier rates"
)

population_sa2_model_ready <- population_sa2_clean %>%
  mutate(across(all_of(key_cols_pop), winsorise_iqr, .names = "{.col}_winsor"))

income_sa2_model_ready <- income_sa2_clean %>%
  mutate(across(all_of(key_cols_inc), winsorise_iqr, .names = "{.col}_winsor"))
```

Although the data file nominally covers 2011-2025, not all variables are equally complete every year. The next figure checks whether the key variables are concentrated in a specific year, which will inform future modeling whether the range of years should be limited.

```{r}
# ── Overall non-missing value % by year in SA2 ─────────────────────────────
id_cols_full <- c("Code", "Label", "Year", "source", "t2", "t3", "geo_level")

key_cols_pop_tmp <- names(population_sa2)[str_detect(str_to_lower(names(population_sa2)), "total") & !(names(population_sa2) %in% id_cols_full)]
key_cols_inc_tmp <- names(income_sa2)[str_detect(str_to_lower(names(income_sa2)), "total") & !(names(income_sa2) %in% id_cols_full)]

pop_year_nonmiss <- population_sa2 %>%
  group_by(Year) %>%
  summarise(non_missing_rate = mean(!is.na(as.matrix(across(all_of(key_cols_pop_tmp))))), .groups = "drop") %>%
  mutate(file = "Population")

inc_year_nonmiss <- income_sa2 %>%
  group_by(Year) %>%
  summarise(non_missing_rate = mean(!is.na(as.matrix(across(all_of(key_cols_inc_tmp))))), .groups = "drop") %>%
  mutate(file = "Income")

year_nonmiss_df <- bind_rows(pop_year_nonmiss, inc_year_nonmiss)

ggplot(year_nonmiss_df, aes(x = Year, y = non_missing_rate, color = file)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ file, ncol = 2) +
  scale_y_continuous(labels = percent_format()) +
  scale_color_manual(values = c("Population" = "steelblue", "Income" = "coral")) +
  labs(
    title = "SA2 overall completeness over time",
    x = "Year",
    y = "Non-missing rate"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The following table gives a more detailed view: each key column is checked by year. If the coverage of a year in key columns is very weak, care should be taken in modeling, even if some rows still exist.

```{r, fig.width=11, fig.height=6}
key_year_completeness <- bind_rows(
  population_sa2 %>%
    group_by(Year) %>%
    summarise(across(all_of(key_cols_pop_tmp), ~ mean(!is.na(.x))), .groups = "drop") %>%
    pivot_longer(-Year, names_to = "variable", values_to = "non_missing_rate") %>%
    mutate(file = "Population"),
  income_sa2 %>%
    group_by(Year) %>%
    summarise(across(all_of(key_cols_inc_tmp), ~ mean(!is.na(.x))), .groups = "drop") %>%
    pivot_longer(-Year, names_to = "variable", values_to = "non_missing_rate") %>%
    mutate(file = "Income")
)

ggplot(
  key_year_completeness,
  aes(x = factor(Year), y = str_trunc(variable, 45), fill = non_missing_rate)
) +
  geom_tile(color = "white") +
  facet_wrap(~ file, scales = "free_y") +
  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  scale_fill_gradient(low = "grey90", high = "steelblue", labels = percent_format()) +
  labs(
    title = "Key-variable completeness by year",
    x = "Year",
    y = NULL,
    fill = "Complete"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),
    panel.grid = element_blank()
  )
```

As a data integrity check, we verified that the total sum of the percentage decomposition columns is about 100 per row. The sum of the age group percentage column (for individuals) and the income level percentage column should be 100. Rows far beyond this range represent remaining data quality issues.

```{r}
# ── Sum check: percentage columns should sum to ~100 per row ───────────────
age_pct_cols <- names(population_sa2_clean)[
  str_detect(names(population_sa2_clean), "Persons") &
    str_detect(names(population_sa2_clean), "years") &
    str_detect(names(population_sa2_clean), fixed("(%)"))
]

inc_pct_cols <- names(income_sa2_clean)[
  str_detect(str_to_lower(names(income_sa2_clean)), "per week \\(%\\)")
]

cat("── Age group % sum check (Persons, should be ~100) ──────\n")
if (length(age_pct_cols) > 0) {
  s <- rowSums(population_sa2_clean[, age_pct_cols, drop = FALSE])
  cat("  Columns :", length(age_pct_cols), "\n")
  cat("  Mean    :", sprintf("%.2f", mean(s)), "   Min :", sprintf("%.2f", min(s)), "   Max :", sprintf("%.2f", max(s)), "\n")
  cat("  Rows outside [95, 105] :", format(sum(s < 95 | s > 105), big.mark = ","), "\n")
} else {
  cat("  No matching columns found\n")
}

cat("\n── Income bracket % sum check (should be ~100) ──────────\n")
if (length(inc_pct_cols) > 0) {
  s2 <- rowSums(income_sa2_clean[, inc_pct_cols, drop = FALSE])
  cat("  Columns :", length(inc_pct_cols), "\n")
  cat("  Mean    :", sprintf("%.2f", mean(s2)), "   Min :", sprintf("%.2f", min(s2)), "   Max :", sprintf("%.2f", max(s2)), "\n")
  cat("  Rows outside [95, 105] :", format(sum(s2 < 95 | s2 > 105), big.mark = ","), "\n")
} else {
  cat("  No matching columns found\n")
}
```

### 3-2 Bivariate Analysis

Since the target variable does not exist in the raw data, we engineer it here. The analysis dataset is built with a **full outer join** at SA2-year level, so valid rows from either file are not discarded too early. We then add year-on-year change variables and classify each SA2-year observation by two dimensions:

- **Age trajectory**: ageing, stable, or growing younger, based on the year-on-year change in median age.
- **Population movement**: population increase, stable population, or population decrease, based on the year-on-year percentage change in estimated resident population.

This creates both the required three-class classification target and a more detailed 3 x 3 exploratory label.

| Label | Rule |
|---|---|
| `aging` | Median age increased by > 0.5 years YoY |
| `stable` | Change within +/- 0.5 years |
| `growing_younger` | Median age decreased by > 0.5 years YoY |

For population movement, the threshold is +/-1% year-on-year population change. This avoids treating small annual fluctuations as meaningful movement.

```{r}
# ---- Build full-join analysis data with short names ----
id_cols_full <- c("Code", "Label", "Year", "source", "t2", "t3", "geo_level")

make_short_named_df <- function(df, prefix) {
  id_df <- df %>%
    transmute(
      code = str_trim(as.character(Code)),
      region = Label,
      year = as.integer(Year)
    )

  data_df <- df %>%
    select(-any_of(id_cols_full))

  short_names <- make.unique(shorten_abs_name(names(data_df)), sep = "_dup")
  names(data_df) <- paste0(prefix, short_names)

  bind_cols(id_df, data_df)
}

pop_for_join <- make_short_named_df(population_sa2_model_ready, "pop_")
inc_for_join <- make_short_named_df(income_sa2_model_ready, "inc_") %>%
  select(-region)

analysis_df <- full_join(pop_for_join, inc_for_join, by = c("code", "year")) %>%
  arrange(code, year)

cat("Full-joined analysis dataset:", format(nrow(analysis_df), big.mark = ","), "rows x", ncol(analysis_df), "columns\n")

median_age_var <- names(analysis_df)[str_detect(names(analysis_df), "^pop_median_age_person_yrs$")][1]
erp_var <- names(analysis_df)[str_detect(names(analysis_df), "^pop_erp_n$")][1]

analysis_df <- analysis_df %>%
  group_by(code) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    median_age_yoy = .data[[median_age_var]] - lag(.data[[median_age_var]]),
    erp_yoy_pct = (.data[[erp_var]] - lag(.data[[erp_var]])) / lag(.data[[erp_var]])
  ) %>%
  ungroup() %>%
  mutate(
    age_trajectory = case_when(
      is.na(median_age_yoy) ~ NA_character_,
      median_age_yoy > 0.5 ~ "aging",
      median_age_yoy < -0.5 ~ "growing_younger",
      TRUE ~ "stable"
    ),
    population_movement = case_when(
      is.na(erp_yoy_pct) ~ NA_character_,
      erp_yoy_pct > 0.01 ~ "population_increase",
      erp_yoy_pct < -0.01 ~ "population_decrease",
      TRUE ~ "stable_population"
    ),
    trajectory_9class = if_else(
      is.na(age_trajectory) | is.na(population_movement),
      NA_character_,
      paste(age_trajectory, population_movement, sep = "__")
    )
  )

combined_labelled <- analysis_df %>%
  filter(!is.na(age_trajectory))

trajectory_tbl <- combined_labelled %>%
  count(age_trajectory, name = "rows") %>%
  mutate(share = percent(rows / sum(rows), accuracy = 0.1))

movement_tbl <- combined_labelled %>%
  filter(!is.na(trajectory_9class)) %>%
  count(age_trajectory, population_movement, name = "rows") %>%
  group_by(age_trajectory) %>%
  mutate(within_age_share = percent(rows / sum(rows), accuracy = 0.1)) %>%
  ungroup()

knitr::kable(trajectory_tbl)

movement_display_tbl <- movement_tbl %>%
  mutate(
    age_trajectory = str_to_title(str_replace_all(age_trajectory, "_", " ")),
    population_movement = str_to_title(str_replace_all(population_movement, "_", " "))
  ) %>%
  rename(
    `Age trajectory` = age_trajectory,
    `Population movement` = population_movement,
    Rows = rows,
    `Within age group` = within_age_share
  )

knitr::kable(
  movement_display_tbl,
  caption = "Population movement within each age trajectory class",
  col.names = c("Age trajectory", "Population movement", "Rows", "Within age group")
)
```

Next, we add the selected annual change characteristics. These columns help later classifiers because the results are about changes, not just the current level of the region.

```{r}
# ---- Add year-on-year change features for interpretable predictors ----
candidate_change_patterns <- c(
  "pop_pop_density",
  "pop_working_age_pop.*pct",
  "inc_median_eq_hh_income",
  "inc_v_1_499_per_week_pct",
  "inc_v_3000_or_more_per_week_pct"
)

change_feature_cols <- unique(unlist(map(
  candidate_change_patterns,
  ~ names(analysis_df)[str_detect(names(analysis_df), .x)][1]
)))
change_feature_cols <- change_feature_cols[!is.na(change_feature_cols)]

analysis_df <- analysis_df %>%
  group_by(code) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(across(all_of(change_feature_cols), ~ .x - lag(.x), .names = "{.col}_yoy_change")) %>%
  ungroup()

change_summary <- analysis_df %>%
  summarise(across(ends_with("_yoy_change"), ~ mean(!is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "change_feature", values_to = "available_rate") %>%
  mutate(available_rate = percent(available_rate, accuracy = 0.1))

knitr::kable(change_summary)
```

The relationship between features is studied by using correlation heat map. Highly correlated predictions are not necessarily wrong, but they warn us that future models may contain redundant information.

```{r}
# ---- Feature-feature relationships ----
core_feature_cols <- c(
  names(analysis_df)[str_detect(names(analysis_df), "pop_pop_density")][1],
  names(analysis_df)[str_detect(names(analysis_df), "pop_working_age_pop.*pct")][1],
  names(analysis_df)[str_detect(names(analysis_df), "inc_median_eq_hh_income")][1],
  names(analysis_df)[str_detect(names(analysis_df), "inc_v_1_499_per_week_pct")][1],
  names(analysis_df)[str_detect(names(analysis_df), "inc_v_3000_or_more_per_week_pct")][1],
  "median_age_yoy",
  "erp_yoy_pct"
)
core_feature_cols <- unique(core_feature_cols[!is.na(core_feature_cols) & core_feature_cols %in% names(analysis_df)])

cor_df <- analysis_df %>%
  select(all_of(core_feature_cols)) %>%
  select(where(~ {
    non_missing_n <- sum(!is.na(.x))
    is.numeric(.x) && non_missing_n >= 2 && isTRUE(sd(.x, na.rm = TRUE) > 0)
  }))

cor_mat <- cor(cor_df, use = "pairwise.complete.obs")

cor_long <- as.data.frame(as.table(cor_mat)) %>%
  as_tibble() %>%
  rename(feature_1 = Var1, feature_2 = Var2, correlation = Freq)

ggplot(cor_long, aes(x = feature_1, y = feature_2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "coral", mid = "white", high = "steelblue", limits = c(-1, 1)) +
  labs(
    title = "Correlation between selected features",
    x = NULL,
    y = NULL,
    fill = "r"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Finally, we examine the feature result relationship. We use Kruskal Wallis statistics to rank the candidate numerical variables for the trajectories of three age groups, and then use the box graph to display the strongest variables.

```{r}
# ---- Feature-to-outcome relationships ----
safe_kw_stat <- function(x, g) {
  keep <- !is.na(x) & !is.na(g)
  if (sum(keep) < 10 || length(unique(g[keep])) < 2 || sd(x[keep], na.rm = TRUE) == 0) return(NA_real_)
  as.numeric(kruskal.test(x[keep] ~ g[keep])$statistic)
}

predictor_cols <- setdiff(
  names(combined_labelled)[map_lgl(combined_labelled, is.numeric)],
  c("year", "median_age_yoy", "erp_yoy_pct")
)

feature_rank_tbl <- tibble(
  feature = predictor_cols,
  kw_statistic = map_dbl(predictor_cols, ~ safe_kw_stat(combined_labelled[[.x]], combined_labelled$age_trajectory))
) %>%
  filter(!is.na(kw_statistic)) %>%
  arrange(desc(kw_statistic))

knitr::kable(head(feature_rank_tbl, 10))

top_plot_features <- head(feature_rank_tbl$feature, 6)

feature_outcome_plot_df <- combined_labelled %>%
  mutate(age_trajectory = factor(age_trajectory, levels = c("aging", "stable", "growing_younger"))) %>%
  select(age_trajectory, all_of(top_plot_features)) %>%
  pivot_longer(cols = all_of(top_plot_features), names_to = "feature", values_to = "value")

ggplot(feature_outcome_plot_df, aes(x = age_trajectory, y = value, fill = age_trajectory)) +
  geom_boxplot(outlier.alpha = 0.15) +
  facet_wrap(~ feature, scales = "free_y") +
  scale_fill_manual(values = c("aging" = "coral", "stable" = "steelblue", "growing_younger" = "mediumseagreen")) +
  labs(
    title = "Top numeric features by age trajectory",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1), legend.position = "none")
```

At this stage of exploration, the strongest predictor is the variable distributed among the trackgroups. These variables should be given priority in future preprocessing and modeling plans, while highly relevant variables may require regularization, tree based model or feature selection.

### 3-3 Data Visualisation

This section collects other charts that directly support modeling decisions: class imbalances, time trends, and detailed 3 x 3 trajectory patterns.

```{r}
# ---- Outcome class imbalance ----
ggplot(trajectory_tbl, aes(x = reorder(age_trajectory, rows), y = rows, fill = age_trajectory)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = c("aging" = "coral", "stable" = "steelblue", "growing_younger" = "mediumseagreen")) +
  labs(
    title = "Class distribution for the three-class outcome",
    x = NULL,
    y = "Number of SA2-year observations"
  ) +
  theme_minimal()

# ---- Median annual age change by year and class ----
age_trend_tbl <- combined_labelled %>%
  group_by(year, age_trajectory) %>%
  summarise(
    median_yoy_age_change = median(median_age_yoy, na.rm = TRUE),
    rows = n(),
    .groups = "drop"
  )

ggplot(age_trend_tbl, aes(x = year, y = median_yoy_age_change, color = age_trajectory)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_line() +
  geom_point() +
  scale_color_manual(values = c("aging" = "coral", "stable" = "steelblue", "growing_younger" = "mediumseagreen")) +
  labs(
    title = "Median year-on-year age change by trajectory class",
    x = "Year",
    y = "Median age change",
    color = "Class"
  ) +
  theme_minimal()

# ---- Detailed 3 x 3 label heatmap ----
ggplot(movement_tbl, aes(x = population_movement, y = age_trajectory, fill = rows)) +
  geom_tile(color = "white") +
  geom_text(aes(label = rows), size = 3) +
  scale_fill_gradient(low = "grey90", high = "steelblue") +
  labs(
    title = "Detailed 3 x 3 demographic trajectory pattern",
    x = "Population movement",
    y = "Age trajectory",
    fill = "Rows"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

eda_insight_tbl <- tibble(
  EDA_area = c("Missing values", "Outliers", "Feature-feature", "Feature-outcome", "Class balance"),
  Modelling_implication = c(
    "Use key-column filtering first; keep and transparently impute/carry NA for less central breakdown variables.",
    "Cap key numeric predictors for model-ready copies, because many extreme SA2 values are plausible rather than data errors.",
    "Check correlated predictors before modelling; regularised or tree-based models can handle redundancy better than distance-based models.",
    "Prioritise top-ranked variables from the Kruskal-Wallis table for initial feature selection.",
    "Use stratified train/test splitting and class-sensitive metrics such as macro-F1 or balanced accuracy."
  )
)

knitr::kable(eda_insight_tbl)
```
